### 2012-09-19 paper/chemokine_main.tex
Paper Version submitted to PLoS CB

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung quantifies constraints of chemokine directed migration}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$
Candice Clay$^{2}$, 
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$, 
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci comprising a tiny fraction of the total lung.  T cell navigation to the infected region is undirected but focused migration is made possible by local cytokine and chemokine production from infected epithelial cells.  We examined the efficiency of local chemokines to induce migration of activated CD8 T cells using simulations in a purpose-built agent-based mathematical model (CyCells).

Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.  Of the chemokines tested, only secretion of CXCL10 (IP-10) and CCL5 (RANTES) was stimulated by infection with these three influenza strains, with IP-10 dominating the directed migration due to higher concentrations.  A delayed differential equation model was fit to the empirical per cell chemokine production rates, coupled with published T cell migration parameters, to calibrate the spatially explicit model in order to test inter-strain variation on T cell recruitment in the lungs.  We observe that T cell sensitivity to chemokine is just high enough to maximize their efficiency.  The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production and large plaque size.

The spatial nature of the model reveals unique challenges to T cell recruitment and motility not visible in spatially homogeneous ODE models.  Expanding plaque sizes isolate infected cells, impeding T cell discovery and infection control. In the model, T cells executed an efficient search with high sensitivity to either chemokine detected. A key limitation imposed on T cells was illustrated in their failure to clear the most rapidly replicating pandemic H1N1 virus after day 6, when T cells became ‘lost’ and inefficient in large infected foci as revealed in cinematic displays. This spatially explicit model is consistent with an efficient T cell recruitment to infected lung, without apparent means to induce increased efficiency, leaving vaccination as the key to shortening the interval between initial infection and T cell activation.

@@

Seasonal influenza infects up to 10,000,000 people and kills between 250,000 and 500,00 people every year.  Recent mutant strains of flu with high morbidity such as Avian H5N1 and 2009 Pandemic H1N1 raise the possibility of a global pandemic in the near future.  In order to clear an infection in the lung, T cells produced in the lymph nodes must navigate the branching bronchial vascular system in order to reach the site of infection.  Using an agent-based model, we look at how molecular signaling by cells infected with three differing strains of influenza aid this search process.  Our experimental data shows that two chemokines, IP-10 and RANTES, are help improve T cell search.  The spatial nature of our model reveals challenges to this search process not seen in conventional ordinary differential equation (ODE) models.  The simulated immune response is not able to contain the highly virulent 2009 pandemic H1N1 strain, suggesting that early intervention, i.e. vaccination, is key to containing the infection.

@@

To characterize the local chemokine-secretion environment in which T cells migrate, we used data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains of avian, seasonal and pandemic origins \cite{Mitchell2011}.   We selected in vitro chemokine data as a tissue environment surrogate, because it may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and it allowed us to represent sustained chemokine gradients within the model lung.  Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to calibrate the ABM model. [REF]

@@

Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing of infections.  Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T cells to the infected cells.  Although virus is produced at a higher rate than chemokine, its larger size diffuses much more slowly, while chemokine decays more quickly.  These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}). Until day 4 the plaque is dominated by active (incubating and secreting) cells, whereas dead cells are rare. Over time, cells in the plaque's interior die, and active cells form a decreasing proportion of the plaque. T cells arrive at day 5 and begin killing the virus-secreting cells. By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.  In aH5N1, the plaque is dense, allowing T cells to find secreting cells easily, and infection is eliminated.  However, in both H1N1 simulations secreting cells were not eliminated.  Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..  T cells still accumulate, but they arrive more slowly than the plaque is growing, leading to lower average T cell killing rate.  Further, the regions of concentrated chemokine lag behind the cell and virus spatial layout.  It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.  Thus, T cells can fail to detect cellular changes in the plaque.  Taken together, the delayed response and low proportion of virus secreting cells prevent T cells from completely clearing the infection.

T cells are unable to kill cells that have not yet presented antigen.  At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}). By day 7, this ratio is approximately 1:100 for both the pH1N1 and sH1N1 strains.  The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T cells can control (but not eliminate) the sH1N1 infection.

@@

Hypercytokinemia, or 'cytokine storm' is described in virulent influenza infections \cite{DeJong2006} but bloodstream cytokine and chemokine levels may not reflect local tissue concentrations.  The use of in vitro human bronchial epithelial cells \cite{Mitchell2011} infected at low cell:virus infection ratio (MOI) was intended to mimic initial natural infection and subsequent local infection spread.  MOI and degree of differentiation of NHBE cells are critical determinants of cytokine/chemokine secretion \cite{Chan2010}.   Yet, as we noted at low MOI, greater RANTES secretion by H5N1 strains has been found in NHBE cultures infected at high MOI with seasonal \cite{Chan2005, Chan2010, Zeng2011} and pandemic \cite{Zeng2011} strains.  Thus chemokine secretion stimulated by H5N1 strains does not appear to be attenuated in parallel with the type 1 interferon response \cite{Zeng2007}, placing greater importance on the role of T cells in controlling the infection.

@@

Can T cell recruitment to tissue during the primary immune response be manipulated to increase search efficiency, with particular implications for recovery of infants prior vaccination?  The high sensitivity to chemokine signals and lack of increased efficiency with an added chemokine do not appear to be avenues for manipulation. Increasing inflammatory cells may raise the volume of tissue containing chemokine signal, thus amplifying T cell search success, but comes with the price of greater morbidity. T cell recruitment to regional lymph nodes is highly efficient, and numbers of responding clones is restricted by immunological dominance. We conclude that, with respect to the elements of recruitment studied in this spatially explicit model, recruitment is efficient and offers few if any options for manipulating its efficiency. Only vaccination provides the clonal expansion and tissue resident memory cells to accelerate the appearance of activated T cells in tissue.

@@

We thank Judy Cannon and Francois Asperti-Boursin (University of New Mexico, Department of Pathology) and Sheldon Jordan (University of New Mexico, Center for Biomedical Engineering) for discussions and suggestions concerning cellular and cytokine/chemokine behavior.  We thank George Bezerra, Ben Edwards, and George Stelle (University of New Mexico, Department of Computer Science) for discussions relating to the design and parameterization of the agent-based model.

This publication was funded by NIH grants R21-AI-73607 (to F.K.) and U01-AI-074561 (to F.K.) and by NSF grant XXX (to S.F and M.M.).

>>>>

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung quantifies constraints of chemokine directed migration}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$
Candice Clay$^{2}$, 
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells from regional lymph nodes to infected foci comprising a fraction of the total lung.  T cell navigation to this region is undirected but migration is made possible by local cytokine and chemokine production from infected epithelial cells.  We examined the efficiency of local chemokines to induce migration of activated CD8 T cells using an agent-based model (CyCells).

Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used to induce chemokine production \textit{in vitro}.  Of the chemokines tested, secretion of CXCL10 (IP-10) and CCL5 (RANTES) were stimulated by infection, with IP-10 dominating the migration due to higher concentrations.  A differential equation model was fit to empirical chemokine production rates and coupled with published T cell parameters to calibrate the spatial model to test inter-strain variation on T cell recruitment in the lungs.  We observe that T cell sensitivity to chemokine is high enough to maximize their efficiency.  The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production.

The spatial nature of the model reveals unique challenges to T cell recruitment not visible in spatially homogeneous ODE models.  Infected cells can become isolated in expanding plaques, impeding infection control.  In the model, T cells executed an efficient search with sensitivity to either chemokine.  A key limitation imposed on T cells was illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in large infected foci as revealed in cinematics. This spatial model is consistent with efficient T cell recruitment to infected lung.  Without an apparent means to induce increased efficiency, vaccination remains the key to shortening the interval between initial infection and T cell activation.

@@

Clearance of influenza from the lung depends strongly on the efficiency with which T cells travel from the lymph node to the site of infection.  However, T cells in the lung must navigate the large branching bronchial network.  This maze-like structure is a challenge for T cell search, and the vast majority of T cells exit the lung without encountering infection.  Infected cells produce cytokines and chemokines, which signal T cells to exit the vasculature and enter tissue.  We developed an agent-based model to quantify how these molecular signals affect search.  Our experimental data show that two chemokines, IP-10 and RANTES, are important for T cell search.  We incorporated the data in our spatially explicit model, which allowed us to identify challenges to the search process not obvious in conventional differential equation models.  We ran three sets of simulations, using parameters from seasonal influenza, avian influenza, and 2009 pandemic influenza.  The simulated immune response contains the first two infections but is unable to contain the highly virulent pandemic strain due to the rapidly expanding plaque size.  Direct interventions to improve this T cell search process are unlikely to be effective, and therefore vaccination is required to contain more virulent strains.

@@

To characterize the local chemokine-secretion environment in which T cells migrate, we used data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains of avian, seasonal and pandemic origins \cite{Mitchell2011}.   We selected in vitro chemokine data as a tissue environment surrogate, because it may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and it allowed us to represent sustained chemokine gradients in our model.  Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to calibrate the ABM model \cite{Miao2010, Ganusov2011}.

@@

Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing of infections.  Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T cells to the infected cells.  Although virus is produced at a higher rate than chemokine, its larger size diffuses much more slowly, while chemokine decays more quickly.  These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}). Until day 4 the plaque is dominated by active (incubating and secreting) cells, whereas dead cells are rare. Over time, cells in the plaque's interior die, and active cells form a decreasing proportion of the plaque. T cells arrive at day 5 and begin killing the virus-secreting cells. By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.  Refer to the videos in Figures S3-S5 to see a representative simulation of seasonal influenza.

In aH5N1, the plaque is dense, allowing T cells to find secreting cells easily, and infection is eliminated.  However, secreting cells were not eliminated in either H1N1 simulations.  Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..  T cells still accumulate, but they arrive more slowly than the plaque is growing, leading to lower average T cell killing rate.  Further, the regions of concentrated chemokine lag behind the cell and virus spatial layout.  It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.  Thus, T cells can fail to detect cellular changes in the plaque.  Taken together, the delayed response and low proportion of virus secreting cells prevent T cells from completely clearing the infection.

T cells are unable to kill cells that have not yet presented antigen.  At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}). By day 7, this ratio is approximately 1:100 for both the pH1N1 and sH1N1 strains.  The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T cells can control (but not eliminate) the sH1N1 infection.  The video in Figure S6 illustrates this event.

@@

Hypercytokinemia, or `cytokine storm' is described in virulent influenza infections \cite{DeJong2006} but bloodstream cytokine and chemokine levels may not reflect local tissue concentrations.  The use of in vitro human bronchial epithelial cells \cite{Mitchell2011} infected at low cell:virus infection ratio (MOI) was intended to mimic initial natural infection and subsequent local infection spread.  MOI and degree of differentiation of NHBE cells are critical determinants of cytokine/chemokine secretion \cite{Chan2010}.   Yet, as we noted at low MOI, greater RANTES secretion by H5N1 strains has been found in NHBE cultures infected at high MOI with seasonal \cite{Chan2005, Chan2010, Zeng2011} and pandemic \cite{Zeng2011} strains.  Thus chemokine secretion stimulated by H5N1 strains does not appear to be attenuated in parallel with the type 1 interferon response \cite{Zeng2007}, placing greater importance on the role of T cells in controlling the infection.

@@

Can T cell recruitment to tissue during the primary immune response be manipulated to increase search efficiency, with particular implications for recovery of people prior vaccination?  The high sensitivity to chemokine signals and lack of increased efficiency with an added chemokine do not appear to be avenues for manipulation. Increasing inflammatory cells may raise the volume of tissue containing chemokine signal, thus amplifying T cell search success, but comes with the price of greater morbidity. T cell recruitment to regional lymph nodes is highly efficient, and numbers of responding clones is restricted by immunological dominance. We conclude that, with respect to the elements of recruitment studied in this spatially explicit model, recruitment is efficient and offers few if any options for manipulating its efficiency. Only vaccination provides the clonal expansion and tissue resident memory cells to accelerate the appearance of activated T cells in tissue.

@@

We thank J. Cannon (UNM Pathology and Molecular Genetics \& Microbiology), F. Asperti-Boursin (UNM Pathology and Computer Science), S. Jordan (UNM Computer Science and Biology), and G. Bezerra, B. Edwards, and G. Stelle (UNM Computer Science) for suggestions concerning the manuscript, cellular and chemokine behavior, and agent-based modeling.

This publication was funded by NIH grants R21-AI-73607 (to F.K.) and U01-AI-074561 (to F.K.), DARPA grant P-1070-113237 (to S.F and M.M), and NSF grant EF-1038682 (to S.F and M.M.).

### 2012-09-19 paper/chemokine_supplement.tex
Paper Version submitted to PLoS CB

>>>>

\section*{Video}

\begin{figure}
\caption{First of three overlaid videos of a representative seasonal H1N1 infection.  This video spans the 10 day infection and shows the cells as they transition from healthy to infected to dead.  T cells show half way through the simulation.  Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.} 
 \label{video:cell_view}
\end{figure}

\begin{figure}
\caption{Second of three overlaid videos of a representative seasonal H1N1 infection.  This video spans the 10 day infection and shows the virus concentration.  Notice the volatility when T cells arrive halfway through the simulation.  Virus concentration ranges from 1e-13 mols/mL (white) to 1e-27 mols/mL (black).  Refer to Figure 5 for the detailed legend. } 
 \label{video:virus_view}
\end{figure}

\begin{figure}
\caption{Third of three overlaid videos of a representative seasonal H1N1 infection.  This video spans the 10 day infection and shows the chemokine concentration.  Notice the volatility when T cells arrive halfway through the simulation.  Chemokine concentration ranges from 1e8 ng/mL (white) to 1e-6 ng/mL (black).  Refer to Figure 5 for the detailed legend. } 
 \label{video:chemokine_view}
\end{figure}

\begin{figure}
\caption{A closer look at the 2009 pandemic simulation.  This video shows the infection from day 6 to day 7 with each frame spanning 100 simulated seconds.  Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.  Note the high proportion of virus-secreting cells (orange) early on.  As time passes, secreting cells are gradually contained to the point where they become very sparse.  T cell clumping often prevents the T cells from quick discovery of new secreting cells.}
\end{figure}
